The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-07 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
David Kaplan's comments and finishing the author lists

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\Curtin{$^{3}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~Kaplan...,
... , %\newauthor
builders list
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data, and another 0.14\% is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV). After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data. Some advise is given to ensure the SKA is robust to RFI, which includes: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

\section{Introduction}
Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider frequency bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing. It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows the customisation of its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency directions. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Visibility amplitude distributions with logarithmic axes. The dashed line represent fits over a reasonably constant part of the tail (selected by eye) of the RFI distribution: $N=\beta S^{-1.43}$ for GLEAM, $N=\beta S^{-4.47}$ for EoR high and $N=\beta S^{-1.33}$ for EoR low.}
\label{fig:lognlogs}
\end{center}
\end{figure*}

\subsection{Distribution analysis}
Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that observations cover different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI-free (green lines). The EoR high-band curves do not include the night that was contaminated by DTV RFI.

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows a comparison of RFI at the same frequencies between the telescopes. The RFI occupancies for LOFAR from \citep{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citep{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown by \citep{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. \citep{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits, and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

>>>>

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\ASTRON{$^{3}$}
\def\Curtin{$^{4}$}
\def\UWisc{$^{5}$}
\def\UW{$^{6}$}
\def\CASS{$^{7}$}
\def\SKASA{$^{8}$}
\def\ASU{$^{9}$}
\def\USydney{$^{10}$}
\def\Haystack{$^{11}$}
\def\RRI{$^{12}$}
\def\MIT{$^{13}$}
\def\UWA{$^{14}$}
\def\CfA{$^{15}$}
\def\UMelbourne{$^{16}$}
\def\Victoria{$^{17}$}
\def\Tata{$^{18}$}
%\def\UMichigan{$^{19}$}
%\def\NRAO{$^{20}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
R.~B.~Wayth\Curtin$^,$\CAASTRO,
N.~Hurley-Walker\Curtin,
D.~L.~Kaplan\UWisc,
N.~Barry\UW,
A.~P.~Beardsley\UW,
M.~E.~Bell\CASS,
G.~Bernardi\SKASA,
J.~D.~Bowman\ASU, 
F.~Briggs\ANU,
J.~Callingham\USydney,
R.~J.~Cappallo\Haystack, 
P.~Carroll\UW,
%A.~de~Oliveira-Costa
A.~A.~Deshpande\RRI, 
J.~S.~Dillon\MIT,
K.~S.~Dwarakanath\RRI,
A.~Ewall-Wice\MIT,
L.~Feng\MIT,
B.-Q.~For\UWA,
B.~M.~Gaensler\USydney$^,$\CAASTRO, 
L.~J.~Greenhill\CfA,
P.~Hancock\Curtin,
K.~Han-Seek\UMelbourne,
B.~J.~Hazelton\UW,
J.~N.~Hewitt\MIT,
L.~Hindson\Victoria,
D.~C.~Jacobs\ASU, 
M.~Johnston-Hollitt\Victoria,
A.~Kapinska\UWA$^,$\CAASTRO,
D.~L.~Kaplan\UWisc,
P.~Kittiwisit\ASU,
E.~Lenc\USydney$^,$\CAASTRO,
J.~Line\UMelbourne,
A.~Loeb\CfA,
C.~J.~Lonsdale\Haystack, 
B.~McKinley\ANU$^,$\CAASTRO,
S.~R.~McWhirter\Haystack,
D.~A.~Mitchell\CASS$^,$\CAASTRO, 
M.~F.~Morales\UW, 
E.~Morgan\MIT, 
J.~Morgan\Curtin,
A.~R.~Neben\MIT,
D.~Oberoi\Tata, 
S.~M.~Ord\Curtin$^,$\CAASTRO,
B.~Pindor\UMelbourne,
J.~C.~Pober\UW,
T.~Prabu\RRI, 
P.~Procopio\UMelbourne,
J.~Riding\UMelbourne,  
N.~Udaya~Shankar\RRI, 
P.~Sourabh\RRI,
K.~S.~Srivani\RRI,
L.~Staveley-Smith\UWA,
R.~Subrahmanyan\RRI$^,$\CAASTRO, 
I.~S.~Sullivan\UW,
M.~Tegmark\MIT,
N.~Thyagarajan\ASU,
S.~J.~Tingay\Curtin$^,$\CAASTRO, 
C.~M.~Trott\Curtin$^,$\CAASTRO,
R.~B.~Wayth\Curtin$^,$\CAASTRO, 
R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
A.~Williams\Curtin, 
C.~L.~Williams\MIT,
C.~Wu\UWA,
J.~S.~Wyithe\UMelbourne
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town, 7405, South Africa\\
\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
\RRI{}Raman Research Institute, Bangalore 560080, India\\
\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
\UWA{}University of Western Australia, Crawley, WA 6009, Perth, Australia\\
\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
}
\jid{PASA}
\doi{10.1017/pas.\the\year.xxx}
\jyear{\the\year}
%Citations may be made using the natbib commands \citet{},\citep{} etc.
\usepackage[authoryear]{natbib}
\bibpunct{(}{)}{;}{a}{}{,}
\setlength{\bibsep}{0.3mm}

\begin{document}
\begin{abstract}
The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA). We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights. RFI detection removes 1.1\% of the data. RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation. After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands. The results are compared to a previously published LOFAR RFI survey. The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data. We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
\end{abstract}

@@

\section{Introduction}
Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider bandwidths of radio observatories. Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}. The consequence of this approach is that a certain fraction of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed. It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach. An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

@@

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues (e.g. \citealt{mwacs-2014}), which of course is not surprising given its remote location. PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI. In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

@@

\section{Method} \label{ch:method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer. It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}. It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.

@@

To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}. Specific customisations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view. Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency axes. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ qualitatively in one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise. An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently. Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimising reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

@@

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Visibility amplitude distributions with logarithmic axes. The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.43$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
\label{fig:lognlogs}
\end{center}
\end{figure*}

\subsection{Distribution analysis}
Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that observations cover different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI-free (green lines). The EoR high-band curves do not include the night that was contaminated by DTV RFI.

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.

@@

\section{Comparison with LOFAR} \label{ch:lofar-comparison}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows a comparison of RFI at the same frequencies between the telescopes. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citet{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown by \citet{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. \citet{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.

@@

SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI. Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem. Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth. Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.
